Geometrical and physical interpretation of the action principle

### 2022-09-28 Articles/2022-ActionPrinciple/ActionPrincipleArchive.tex
ActionPrinciple: uploaded to arXiv

While the principle of stationary action is regarded by many as one of the most important tools in physics, its physical meaning is not completely clear. First of all, the typical characterization of the Lagrangian as the difference between kinetic and potential energy fails even for simple systems, like a charged particle under a magnetic field. Moreover, the Lagrangian for a system is not uniquely defined, making the actual value of the action for a path not directly physically significant.  We are left to wonder: what exactly is the action and why is it stationary for actual trajectories?

As part of our larger project Assumptions of Physics, we developed an approach, called Reverse Physics (TODO cite paper), which examines current theories to find a set of starting physical assumptions that are sufficient to rederive them. We have found that Lagrangian mechanics is equivalent to three assumptions: determinism/reversibility, independence of degrees of freedom and kinematics/dynamics equivalence (TODO cite long paper). This physically motivated understanding of the classical theory can be used to characterize both the physics and the geometry underlying the principle of stationary action. What we find is that this arises as a general mathematical feature of divergence-free fields (and closed two-forms), which are the appropriate tools to describe a flow that conserves the number of states. The assumption of equivalence between kinematics and dynamics (i.e. we can reconstruct the dynamical state simply by looking at the trajectory) is what allows us to express the principle in the usual form. The argument can proceed in the reverse direction: assuming the principle of stationary action recovers a dynamical system that exhibits those three physical assumptions.

The mathematics needed to run the argument is not new\cite{souriau1997structure}. The two aspects that have been sorely lacking, and that we provide, are (1) a clear geometrical interpretation of the action principle and (2) a tight connection between the math and the physics it represents. To make the result accessible to the widest audience, we first cover the case of a single degree of freedom using standard vector calculus. We then proceed to the case of multiple independent degrees of freedom using tools from differential geometry. The main article will presents all the key point needed to follow the argument and its physical and geometrical meaning, leaving the mathematical details and calculations to the appendix.

@@

As we want to characterize the evolution of states over time, the appropriate setting is phase space extended with the time variable.(TODO find citations) That is, the space charted by position $q$, momentum $p$ and time $t$ as can be seen in Fig. \ref{extended_phase_space}. In the same way that we write $x^i = [ x, y, z ]$ for the three dimensions of space, we write
\begin{equation}\label{sdof_variables}
	\xi^a = [ q, p, t]
\end{equation}
for the three dimensions of the extended phase space.

@@

The flow of $\vec{S}$ will still be divergence-free, but this property only tells us that the total number of states is conserved. We also need to assume that
\begin{align}\label{assum_indep}
	\tag*{(IND)}
	\parbox{2.8in}{\emph{the degrees of freedom are independent.}}
\end{align}
This means that, at each moment in time, the total number of states (i.e. the volume) of a parallelepiped will be the product of states identified by each degree of freedom (i.e. the areas). To be able to quantify the number of states identified by each degree of freedom, we introduce the rank-2 tensor (more precisely a two-form)
\begin{equation}\label{mdof_form}
	\omega = \omega_{ab} \, e^a \otimes e^b,
\end{equation}
which we call the state counting form.\footnote{Note that this approach is consistent with statistical mechanics. (TODO: add references)}

@@

Another interesting element is that a version of the principle of stationary actions holds even if \ref{assum_kineq} does not apply. In this case the ``Lagrangian'' $L(q,\dot{q},p,t)$ depends on conjugate momentum as well. Also note that the cases where the Lagrangian does not admit a Hamiltonian are exactly the cases where the action principle fails to yield a unique solution. In this sense, the action principle is better understood as a feature of Hamiltonian mechanics, instead of Lagrangian mechanics.(TODO: check Arnold's book for reference)

@@

(TODO: Describe the space) The case of multiple degrees of freedom proceeds in the same way as the single degree of freedom. We will have to:
\begin{enumerate}[label=(\roman*)]
	\item show that assuming determinism and reversibility \ref{assum_detrev} and independence of degrees of freedom \ref{assum_indep} is equivalent to assuming the existence of the displacement field $\vec{S}$ defined by \ref{mdof_displacement} and the closed counting form $\omega$ defined by \ref{mdof_form} such that $\omega$ is killed only by the displacement field $\vec{S}$ as in \ref{mdof_displacement_kills} or by a field everywhere tangent to $\vec{S}$.
	\item show that the potential $\theta$ for $\omega$ defined in \ref{mdof_form_potential} can be expressed, without loss of generality, as \ref{mdof_potential_expression} and that $\omega$ can be expressed as $\ref{mdof_form_components}$
	\item show that \ref{mdof_form_potential}, \ref{mdof_displacement_kills} and \ref{mdof_potential_expression} recover Hamilton's equations as in \ref{mdof_Ham_eq}
	\item show that the integral of $\omega$ through a surface $\Sigma$ delimited by a path $\gamma$ and its variation $\gamma'$ is the variation of the integral of the vector potential, as claimed in \ref{mdof_action}
	\item show that the integral of the vector potential is stationary for and only for the solutions of the equations of motion
	\item show that the integral of the vector potential reduces to the action if and only if \ref{assum_kineq} applies.
\end{enumerate}

>>>>

While the principle of stationary action is regarded by many as one of the most important tools in physics,\cite{hamilton1834general, heaviside1903, feynman1942principle, brenier1989least, rojo2018principle} its physical meaning is not completely clear. First of all, the typical characterization of the Lagrangian as the difference between kinetic and potential energy fails even for simple systems, like a charged particle under a magnetic field. Moreover, the Lagrangian for a system is not uniquely defined, making the actual value of the action for a path not directly physically significant.  We are left to wonder: what exactly is the action and why is it stationary for actual trajectories?

As part of our larger project Assumptions of Physics, we developed an approach, called Reverse Physics \cite{aop-phys-ReversePhysics}, which examines current theories to find a set of starting physical assumptions that are sufficient to rederive them. We have found that Lagrangian mechanics is equivalent to three assumptions: determinism/reversibility, independence of degrees of freedom and kinematics/dynamics equivalence \cite{aop-phys-blueprint}. This physically motivated understanding of the classical theory can be used to characterize both the physics and the geometry underlying the principle of stationary action. What we find is that this arises as a general mathematical feature of divergence-free fields (and closed two-forms), which are the appropriate tools to describe a flow that conserves the number of states. The assumption of equivalence between kinematics and dynamics (i.e. we can reconstruct the dynamical state simply by looking at the trajectory) is what allows us to express the principle in the usual form. The argument can proceed in the reverse direction: assuming the principle of stationary action recovers a dynamical system that exhibits those three physical assumptions.

The mathematics needed to run the argument is well established\cite{souriau1970structure, abraham1978foundations,arnold1989mathematical, marsden1999introduction}. The two aspects that have been sorely lacking, and that we provide, are (1) a clear geometrical interpretation of the action principle and (2) a tight connection between the math and the physics it represents. To make the result accessible to the widest audience, we first cover the case of a single degree of freedom using standard vector calculus. We then proceed to the case of multiple independent degrees of freedom using tools from differential geometry. The main article will presents all the key point needed to follow the argument and its physical and geometrical meaning, leaving the mathematical details and calculations to the appendix.

@@

As we want to characterize the evolution of states over time, the appropriate setting is phase space extended with the time variable\cite{lanczos1949variational,synge1960encyclopedia}. That is, the space charted by position $q$, momentum $p$ and time $t$ as can be seen in Fig. \ref{extended_phase_space}. In the same way that we write $x^i = [ x, y, z ]$ for the three dimensions of space, we write
\begin{equation}\label{sdof_variables}
	\xi^a = [ q, p, t]
\end{equation}
for the three dimensions of the extended phase space.

@@

The flow of $\vec{S}$ will still be divergence-free, but this property only tells us that the total number of states is conserved. We also need to assume that
\begin{align}\label{assum_indep}
	\tag*{(IND)}
	\parbox{2.8in}{\emph{the degrees of freedom are independent.}}
\end{align}
This means that, at each moment in time, the total number of states (i.e. the volume) of a parallelepiped will be the product of states identified by each degree of freedom (i.e. the areas). To be able to quantify the number of states identified by each degree of freedom, we introduce the rank-2 tensor (more precisely a two-form)
\begin{equation}\label{mdof_form}
	\omega = \omega_{ab} \, e^a \otimes e^b,
\end{equation}
which we call the state counting form.\footnote{Note that this approach is consistent with statistical mechanics\cite{peliti2011statistical}.}

@@

Another interesting element is that a version of the principle of stationary actions holds even if \ref{assum_kineq} does not apply. In this case the ``Lagrangian'' $L(q,\dot{q},p,t)$ depends on conjugate momentum as well. Also note that the cases where the Lagrangian does not admit a Hamiltonian are exactly the cases where the action principle fails to yield a unique solution. In this sense, the action principle is better understood as a feature of Hamiltonian mechanics, instead of Lagrangian mechanics\cite{souriau1970structure, arnold1989mathematical}.

@@

The case of multiple degrees of freedom proceeds in the same way as the single degree of freedom. The main difference is that we will work with a system described by $N$ degrees of freedom, which is therefore described by the $2N+1$ variables
\begin{equation}
\tag*{(\ref{mdof_variables})}
	\xi^a = [ q^i, p_i, t].
\end{equation}
We will have to:
\begin{enumerate}[label=(\roman*)]
	\item show that assuming determinism and reversibility \ref{assum_detrev} and independence of degrees of freedom \ref{assum_indep} is equivalent to assuming the existence of the displacement field $\vec{S}$ defined by \ref{mdof_displacement} and the closed counting form $\omega$ defined by \ref{mdof_form} such that $\omega$ is killed only by the displacement field $\vec{S}$ as in \ref{mdof_displacement_kills} or by a field everywhere tangent to $\vec{S}$.
	\item show that the potential $\theta$ for $\omega$ defined in \ref{mdof_form_potential} can be expressed, without loss of generality, as \ref{mdof_potential_expression} and that $\omega$ can be expressed as $\ref{mdof_form_components}$
	\item show that \ref{mdof_form_potential}, \ref{mdof_displacement_kills} and \ref{mdof_potential_expression} recover Hamilton's equations as in \ref{mdof_Ham_eq}
	\item show that the integral of $\omega$ through a surface $\Sigma$ delimited by a path $\gamma$ and its variation $\gamma'$ is the variation of the integral of the vector potential, as claimed in \ref{mdof_action}
	\item show that the integral of the vector potential is stationary for and only for the solutions of the equations of motion
	\item show that the integral of the vector potential reduces to the action if and only if \ref{assum_kineq} applies.
\end{enumerate}
